Arms Transfers and Frozen Conflict Dynamics

### 2024-01-23 paper.qmd
Manuscript edits

@@

```{R, engine = 'tikz'}
#| label: fig-states
#| fig-cap: State diagram showing the hidden states and their possible
#|          transition paths for a single conflict-year. Note,
#|          "Terminated" is *not* explicitly modeled in the HMM being
#|          presented in this paper. It is shown in order to emphasize
#|          that when a civil conflict is conclusive terminated, it
#|          drops out of the dataset. In other words, "Terminated", is
#|          an absorbing state with no possibility of transitioning to
#|          "Active" or "Frozen".

@@

\end{tikzpicture}

@@

Since $y_t \in \mathbb{N}_0$, I model the yearly battle related deaths given that $S_t$ is in state $k$ with a negative binomial distribution using the mean-dispersion parameterization,

@@

To complete the full inference model I set the following priors. For the initial state probabilities and the baseline intercepts in the time-varying transition function $\alpha_{i,jk}$:

\begin{align*}
\pi & \sim \text{Dirichlet}\left(\frac{1}{K}, \ldots, \frac{1}{K}\right) \\
\nu_{i,jk} & \sim \text{Normal}(\mu_{jk}, \sigma_{jk}) \\
\mu_{jk} & \sim \text{Normal}(0, 1) \\
\sigma_{jk} & \sim \text{Cauchy}_+(0, 0.5)
\end{align*}

Thus, the varying intercepts, $\nu_{\cdot,jk}$, are given hierarchical priors in order to partially pool information across conflicts. Meanwhile, for each regression coefficient, $\beta_d \in \mathbb{R}^{K \times K}$ for $d = 1, 2, \ldots, D$ covariates, I place an uninformative multivariate normal prior with an LKJ prior [@lewandowski_generating_2009] on the correlation matrix, $\Omega$, and a standard Half-Cauchy on the diagonal of variances, $\upsilon$.

@@

\begin{align*}
\phi & \sim \text{Gamma}(2, 0.1) \\
\mu_{ij} & \sim \text{LogNormal}\left(\lambda_j, \tau_j \right) \\
\lambda_1 & \sim \text{Normal}(0, 1) \\
\lambda_2 & \sim \text{Normal}(6.2, 1) \\
\tau & \sim \text{Cauchy}_+(0, 0.5)
\end{align*}

@@

library(dplyr)

@@

dataset <- readRDS("./data/model_data.rds")
fit <- readRDS("./fit.rds")

Z_star <- fit$draws("Z_star", format = "matrix")
active_state <- Z_star[, sprintf("Z_star[%d,2]", 1:nrow(dataset))]

@@

tiv <- fit$draws(c(sprintf("beta[1,%d,2]", 1:2), sprintf("beta[2,%d,2]", 1:2)),
                 format = "matrix")

@@

>>>>

Moving beyond operationalizing conflict episodes as discrete events separated by drops below an arbitrary threshold in the number of battle related fatalities.

@@

```{R, engine = 'tikz'}
#| label: fig-states
#| fig-cap: State diagram showing the hidden states and their possible
#|          transition paths for a single conflict-year. Note,
#|          "Terminated" is *not* explicitly modeled in the HMM being
#|          presented in this paper. It is shown in order to emphasize
#|          that when a civil conflict is conclusively terminated, it
#|          drops out of the dataset. In other words, "Terminated", is
#|          an absorbing state with no possibility of transitioning to
#|          "Active" or "Frozen".

@@

@@

Since $y_t \in \mathbb{N}_0$, I model the yearly battle related deaths given that $S_t$ is in state $k$ with a negative binomial distribution using a mean-dispersion parameterization,

@@

To complete the full inference model I set the following priors. For the initial state probability simplex I select a flat dirichlet prior, $\pi \sim \text{Dirichlet}\left(\frac{1}{K}, \ldots, \frac{1}{K}\right)$.[^dirichlet] For the baseline intercepts in the time-varying transition function, $\alpha_{i,jk}$, I place a hierarchical prior.

\begin{align*}
\nu_{i,jk} & \sim \text{Normal}(\mu_{jk}, \sigma_{jk}) \\
\mu_{jk} & \sim \text{Normal}(0, 1) \\
\sigma_{jk} & \sim \text{Cauchy}_+(0, 0.5)
\end{align*}

[^dirichlet]: TODO: This is an uninformative prior that encodes an equal prior probability for the initial observation to be either a frozen or active. Practically, this will have a relatively small impact on the actual model estimates since the likelihood should dominate the prior; however, conceptually it may be more consistent to give more probability weight to the active state since frozen conflicts by definition should start as active conflicts. The complication is that there is left-censoring in the dataset --- our conflict sequences reflect the start date of the BRD dataset and may not capture the actual first conflict-year coded by UCDP.

Thus, the varying intercepts, $\nu_{\cdot,jk}$, are partially pooled across conflicts. Meanwhile, for each regression coefficient, $\beta_d \in \mathbb{R}^{K \times K}$ for $d = 1, 2, \ldots, D$ covariates, I place an uninformative multivariate normal prior with an LKJ prior [@lewandowski_generating_2009] on the correlation matrix, $\Omega$, and a standard Half-Cauchy on the diagonal of variances, $\upsilon$.

@@

[^prior]: The hierarchical parameters for $\mu_{ij}$ are on the log scale. An informative hyperprior is placed on $\lambda$ where $\mathbb{E}(\lambda) = (0, 6.21)$ corresponding to 1 and ~500 battle related deaths respectively.

\begin{align*}
\phi & \sim \text{Gamma}(2, 0.1) \\
\mu_{ij} & \sim \text{LogNormal}\left(\lambda_j, \tau_j \right) \\
\lambda_1 & \sim \text{Normal}(0, 1) \\
\lambda_2 & \sim \text{Normal}(6.21, 1) \\
\tau & \sim \text{Cauchy}_+(0, 0.5)
\end{align*}

Finally,

@@

library(cmdstanr)
library(dplyr)

@@

dataset <- readRDS("./data/merge_data.rds") |>
    filter(high_intensity)

files <- list.files("posteriors", "csv$", full.names = T)
stopifnot(length(files) > 0)

# This is using data.table::fread which needs R.utils installed to
# read compressed files.
fit <- as_cmdstan_fit(files)

Z_prob <- fit$draws("Z_prob", format = "matrix")
active_state <- Z_prob[, sprintf("Z_prob[%d,2]", 1:nrow(dataset))]

@@

tiv <- fit$draws(c(sprintf("beta[1,%d,1]", 1:2), sprintf("beta[2,%d,1]", 1:2)),
                 format = "matrix")

@@

# Conclusion

Future work:
- incorporate more battlefield dynamics
- dyadic level rather than conflict level
